Search Strategies for Functional Logic Programming

### 2012-01-15 papers/atps2012/paper.tex
ATPS paper: Sect. 2 completed

Curry \cite{Hanus06Curry}
is a declarative multi-paradigm language combining features 
from functional programming (demand-driven evaluation, parametric
polymorphism, higher-order functions) and logic programming
(computing with partial information, unification, constraints).
Recent surveys are available in \cite{AntoyHanus10CACM,Hanus07ICLP}.
The syntax of Curry is close to Haskell \cite{PeytonJones03Haskell}.
Thus, variables and function names usually
start with lowercase letters and the names of type and data constructors
start with an uppercase letter. The application of $f$
to $e$ is denoted by juxtaposition (``$f~e$'').
In addition, Curry allows free (logic) 
variables in conditions and right-hand sides of defining rules.
The operational semantics is based on an optimal evaluation strategy
\cite{AntoyEchahedHanus00JACM} which is a conservative extension
of lazy functional programming and (concurrent) logic programming.

A Curry program consists of the definition of data types and
operations on these types. 
Note that in a functional logic language
operations might yield more than one result on the same input due to 
the logic programming features.
For instance, Curry contains a \emph{choice} operation defined by:
%
\begin{curry}
  x ? _ = x
  _ ? y = y
\end{curry}
%
Thus, the expression \ccode{0$~$?$~$1} has two values: \code{0} and \code{1}.
If expressions have more than one value, one wants to select
intended values according to some constraints,
typically in conditions of program rules.
A \emph{rule} has the form \ccode{$f~t_1\ldots{}t_n$ | $c$ = $e$}
where the (optional) condition $c$ is a \emph{constraint},
i.e., an expression of the built-in type
\code{Success}. For instance, the trivial constraint
\code{success} is a value of type \code{Success} that
denotes the always satisfiable constraint.
Thus, we say that a constraint $c$ is \emph{satisfied} if it
can be evaluated to \code{success}.
An \emph{equational constraint} $e_1 \,\code{=:=}\, e_2$ is satisfiable
if both sides $e_1$ and $e_2$ are reducible to unifiable values.
Furthermore, if $c_1$ and $c_2$ are constraints,
\code{$c_1\,$\&$\,\,c_2$} denotes their concurrent conjunction
(i.e., both argument constraints are concurrently evaluated).

@@

{\it
Mention that search strategy is not fixed
 (discuss top-level vs. encaps. search)}

\section{Search Strategies}
\label{sec:strategies}

@@

\bibliography{mh}
%\bibliography{paper}

>>>>

Integrated functional logic programming languages
combine features 
from functional programming (demand-driven evaluation, parametric
polymorphism, higher-order functions) and logic programming
(computing with partial information, unification, constraints).
Recent surveys are available in \cite{AntoyHanus10CACM,Hanus07ICLP}.
Curry \cite{Hanus06Curry} is a functional logic language
which extends lazy functional programming as to be found in
Haskell \cite{PeytonJones03Haskell} and supports these features.

A Curry program consists of the definition of data types and
operations on these types. 
Since the syntax of Curry is close to Haskell,
variables and function names usually
start with lowercase letters and the names of type and data constructors
start with an uppercase letter. The application of $f$
to $e$ is denoted by juxtaposition (``$f~e$'').
In addition, Curry allows free (logic) 
variables in conditions and right-hand sides of defining rules.
Note that in a functional logic language
operations might yield more than one result on the same input due to 
the logic programming features.
For instance, Curry contains a \emph{choice} operation defined by:
%
\begin{curry}
  x ? _ = x
  _ ? y = y
\end{curry}
%
Thus, the expression \ccode{0$~$?$~$1} has two values: \code{0} and \code{1}.
If expressions have more than one value, one wants to select
intended values according to some constraints,
typically in conditions of program rules.
A \emph{rule} has the form
\begin{curry}
$f~t_1\ldots{}t_n$ | $c$ = $e$
\end{curry}
where the (optional) condition $c$ is a \emph{constraint},
i.e., an expression of the built-in type
\code{Success}. For instance, the trivial constraint
\code{success} is a value of type \code{Success} that
denotes the always satisfiable constraint.
Thus, we say that a constraint $c$ is \emph{satisfied} if it
can be evaluated to \code{success}.
An \emph{equational constraint} $e_1 \,\code{=:=}\, e_2$ is satisfiable
if both sides $e_1$ and $e_2$ are reducible to unifiable values.
%Furthermore, if $c_1$ and $c_2$ are constraints,
%\code{$c_1\,$\&$\,\,c_2$} denotes their concurrent conjunction
%(i.e., both argument constraints are concurrently evaluated).

@@

Curry also offers standard features of
functional languages, like modules or monadic I/O
which is identical to Haskell's I/O concept \cite{Wadler97}.
Thus, \ccode{IO $\alpha$} denotes the type of an I/O action that returns values
of type $\alpha$.

The operational semantics of Curry is based on an optimal evaluation strategy
\cite{AntoyEchahedHanus00JACM} which is a conservative extension
of lazy functional programming and (concurrent) logic programming.
A big-step and a small-step operational semantics of Curry
can be found in \cite{AlbertHanusHuchOliverVidal05}.
Curry's semantics is sound in the sense of logic programming,
i.e., each computed result is correct and for each
correct result there is a more general computed one
\cite{AntoyEchahedHanus00JACM}.
In order to achieve completeness,
one has to take all possible non-deterministic derivation paths
into account. In contrast to Prolog, which fixes
a (potentially incomplete) depth-first search strategy to find solutions,
Curry does not fix a particular search strategy.
Thus, different Curry implementations
can support various search strategies.
For instance, the Curry implementation PAKCS \cite{Hanus10PAKCS},
which compiles Curry programs into Prolog programs,
supports only a depth-first search strategy.
MCC \cite{Lux99FLOPS} compiles Curry programs into C programs
and uses also a depth-first search strategy to find all
solutions of a given top-level goal.
In addition, MCC offers the encapsulation of search
(see below) so that other search strategies,
like a complete breadth-first search strategy, can be used
inside a Curry program.
The Curry implementation KiCS \cite{BrasselHuch07,BrasselHuch09},
which compiles Curry programs into Haskell programs,
offers depth-first and breadth-first search strategies
for top-level goals as well the encapsulation of search
with user-definable strategies.
In this paper we consider the Curry implementation
KiCS2 \cite{BrasselHanusPeemoellerReck11},
which is based on similar ideas than KiCS but uses
a different compilation model avoiding side effects
to enable better optimizations for target programs.
KiCS2 also offers different search strategies for top-level goals
and the encapsulation of search with user-definable strategies,
which is described next.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Search Strategies}
\label{sec:strategies}

As mentioned above, Curry does not enforce a particular search
strategy so that Curry implementation
can provide different search strategies to find solutions
or values for a given constraint or expression.
The most advanced system in this respect is
KiCS2 \cite{BrasselHanusPeemoellerReck11},
which supports the evaluation of top-level expressions
with depth-first search, breadth-first search,
iterative deepening search, or parallel search strategies.
This flexibility is possible since all operations with side effects
are collected in monadic I/O operations \cite{Wadler97}.
As a consequence of this computation model,
all non-deterministic computations between I/O operations
must be encapsulated since one cannot apply two alternative
I/O operations to some world and proceed with two non-deterministic worlds
(``one cannot duplicate the world'').
Therefore, Curry offers the encapsulation of search
by representing non-deterministic computations
in a data structure so that the computation of solutions
(one solution or all solutions) can conceptually be implemented
as traversals on this data structure.

Early: try \cite{HanusSteiner98PLILP,Lux99FLOPS}

Later: \cite{BrasselHanusHuch04JFLP}

@@

%\bibliography{mh}
\bibliography{paper}
